Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-09-18 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
First draft of appendix on the Hodge operator and the semi-implicit method
describe test setup of thermal advection

% Every term in the discretised equations of motion is a potential source of numerical error.
% Errors in calculating the horizontal pressure gradient generate spurious circulations and so much effort has been put into improving this calculation.  This is often achieved by making the grid more orthogonal, flattening terrain following coordinate surfaces [citations] or using a cut cell approach [citations].
% TF citations: mahrer1984, schaer2002, klemp2011
% cut cell citations: good2014, yamazaki2015 puts this clearly but not yet published
% In a test of a stably stratified atmosphere initially at rest, we find that spurious velocities are no more than \SI{0.1}{\meter\per\second} on any grid. [TODO: and this is much better than klemp2011]
% Advection terms also contribute to numerical error.  We show that advection is most accurate when the flow is aligned with the grid layers.  Near the surface, flow tends to follow the terrain which results in larger errors when more orthogonal grids are used.
% In a test of orographically induced gravity waves, we find that the advection of potential temperature generates grid-scale errors on the cut cell grid.

@@

We use the finite volume model from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \mathbf{u}}{\partial t} + \nabla \cdot \rho \mathbf{u}\otimes\mathbf{u} &= \rho \mathbf{g} - c_p \rho \theta \nabla \Pi - \mu \rho \mathbf{u} \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \mathbf{u} &= 0 \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \mathbf{u} \theta &= 0 \label{eq:theta} \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\mathbf{u}\) is the velocity field, \(\mathbf{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.  \(\mu\) is a dimensionless damping function used for the sponge layer in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.

@@

The spatial domain, mountain profile, initial tracer profile and discretisation are the same as those in the horizontal tracer advection test.  The velocity field is defined using a streamfunction, $\Psi$, so that the continuous velocity field is non-divergent and follows the BTF coordinate surfaces given by equation~\ref{eqn:btf} such that
\begin{equation}
	\Psi(x,z) = u_0 H \frac{z - h}{H - h}
\end{equation}
where $u_0 = \SI{10}{\meter\per\second}$, which is the horizontal wind speed where $h(x) = 0$.
The horizontal and vertical components of velocity, $u$ and $w$, are then given by
\begin{align}
	u &= \frac{\partial \Psi}{\partial z} = u_0 \frac{H}{H - h}, \quad w = -\frac{\partial \Psi}{\partial x} = u_0 H \frac{\mathrm{d} h}{\mathrm{d} x} \frac{H - z}{\left( H - h \right)^2}, \nonumber \\
	\frac{\mathrm{d} h}{\mathrm{d} x} &= - h_0 \left[ 
		\beta \cos^2 \left( \alpha x \right) \sin \left( 2 \beta x \right) +
		\alpha \cos^2 \left( \beta x \right) \sin \left( 2 \alpha x \right)
	\right]
\end{align}
Unlike the horizontal advection test, flow extends from the top of the domain all the way to the ground.  The discrete velocity field is calculated using the streamfunction in the same way as the horizontal advection test.

@@

@@

\begin{appendices}
\input{WellerAppendix}
\end{appendices}

%% APPENDICES
%\appendix[A]
%\appendixtitle{Non-orthogonal formulation}

%\TODO{Hilary will redraft this appendix}
%A minor change has been made to the curl-free, non-orthogonal formulation as described in \citet{weller-shahrokhi2014} to improve stability on non-orthogonal grids.  \citet{weller-shahrokhi2014} use a Hodge dual operator, \(H\), to map from co-variant to contra-variant velocity components:
%\begin{align}
%	U = HV
%\end{align}
%where \(U = \rho\mathbf{u}\cdot\mathbf{S}_f\) and \(V=\rho\mathbf{u}\cdot\mathbf{d}_f\).  \(\mathbf{S}_f\) is the surface normal vector for a face, \(f\), which has a magnitude of the surface area, and \(\mathbf{d}_f\) is the vector between neighbouring cell centres.

%In order to create a semi-implicit formulation, \cite{weller-shahrokhi2014} split \(H\) into diagonal and off-diagonal components and the off-diagonal components are treated explicitly in the Helmholtz equation for \(\Pi\).  However, this was found to be unstable when using a large timestep on a grid with large non-orthogonality.  Improved stability is achieved by splitting \(H\) into the orthogonal part and the non-orthogonal correction: 
%\begin{align}
%	H = H_c + H_\mathrm{corr}
%\end{align}
%where the diagonal matrix \(H_c=|\mathbf{S}_f|/|\mathbf{d}_f|\) and the non-orthogonal correction is simply \(H_\mathrm{corr} = H - H_c\).  The orthogonal part, \(H_c\), is treated implicitly in the Helmholtz equation and the non-orthogonal part, \(H_\mathrm{corr}\), is treated explicitly.

>>>>

% Every term in the discretised equations of motion is a potential source of numerical error.
% Errors in calculating the horizontal pressure gradient generate spurious circulations and much effort has been put into improving this calculation.  This is often achieved by making the grid more orthogonal: flattening terrain following coordinate surfaces [citations] or using a cut cell approach [citations].
% TF citations: mahrer1984, schaer2002, klemp2011
% cut cell citations: good2014, yamazaki2015 puts this clearly but not yet published
% In a test of a stably stratified atmosphere initially at rest, we find that spurious velocities are no more than \SI{0.1}{\meter\per\second} on any grid. [TODO: and this is much better than klemp2011]
% Advection terms also contribute to numerical error.  We show that advection is most accurate when the flow is aligned with the grid layers.  Near the surface, flow tends to follow the terrain which results in larger errors when more orthogonal grids are used.
% In a test of orographically induced gravity waves, we find that the advection of potential temperature generates grid-scale errors on the cut cell grid.

@@

We use the finite volume model from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \mathbf{u}}{\partial t} + \nabla \cdot \rho \mathbf{u}\otimes\mathbf{u} &= \rho \mathbf{g} - c_p \rho \theta \nabla \Pi - \mu \rho \mathbf{u} \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \mathbf{u} &= 0 \label{eq:cont} \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \mathbf{u} \theta &= 0 \label{eq:theta} \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0} \label{eq:state}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\mathbf{u}\) is the velocity field, \(\mathbf{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.  \(\mu\) is a dimensionless damping function used for the sponge layer in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.

@@

The spatial domain, mountain profile, initial tracer profile and discretisation are the same as those in the horizontal tracer advection test.  The velocity field is defined using a streamfunction, $\Psi$, so that the continuous velocity field is non-divergent and follows the BTF coordinate surfaces given by equation~\ref{eqn:btf} such that
\begin{equation}
	\Psi(x,z) = u_0 H \frac{z - h}{H - h} \label{eqn:streamfunc-btf}
\end{equation}
where $u_0 = \SI{10}{\meter\per\second}$, which is the horizontal wind speed where $h(x) = 0$.
The horizontal and vertical components of velocity, $u$ and $w$, are then given by
\begin{align}
	u &= \frac{\partial \Psi}{\partial z} = u_0 \frac{H}{H - h}, \quad w = -\frac{\partial \Psi}{\partial x} = u_0 H \frac{\mathrm{d} h}{\mathrm{d} x} \frac{H - z}{\left( H - h \right)^2} \label{eqn:uw-btf} \\
	\frac{\mathrm{d} h}{\mathrm{d} x} &= - h_0 \left[ 
		\beta \cos^2 \left( \alpha x \right) \sin \left( 2 \beta x \right) +
		\alpha \cos^2 \left( \beta x \right) \sin \left( 2 \alpha x \right)
	\right]
\end{align}
Unlike the horizontal advection test, flow extends from the top of the domain all the way to the ground.  The discrete velocity field is calculated using the streamfunction in the same way as the horizontal advection test.

@@

\subsection{Terrain following advection of thermal profile}
The spatial domain, mountain profile, grid resolutions and timesteps are the same as those in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.  The terrain following velocity field is defined by the streamfunction:
\begin{align}
	\Psi(x,z) = u_0 \left\{ \begin{array}{l l}
			H_\mathrm{TF} \frac{z - h}{H_\mathrm{TF} - h} \label{eqn:streamfunc-btf} & \enskip \text{if $z \leq H_\mathrm{TF}$} \\
			z & \enskip \text{if $z > H_\mathrm{TF}$}
	\end{array} \right.
\end{align}
where \(H_\mathrm{TF} = \SI{20}{\kilo\meter}\) is the level at which the terrain following layers become flat; the domain height is \(\SI{30}{\kilo\meter}\).
For $z \leq H$, the $u$ and $w$ components of velocity are given by equation~\ref{eqn:uw-btf}, but $h(x)$ has the same form as equation~\ref{eqn:resting:mountain}, hence the derivative is:
\begin{align}
	\frac{\mathrm{d} h}{\mathrm{d} x} = -h_0 \exp \left(- \left( \frac{x}{a} \right)^2 \right) \left[ \frac{\pi}{\lambda} \sin \left( \frac{2 \pi x}{\lambda} \right) - \frac{2x}{a^2} \cos^2\left( \frac{\pi x}{\lambda} \right) \right]
\end{align}
For $z > H$, $u = u_0$ and $w = 0$.

The tracer field, \(\phi\), is the same as the initial potential temperature field in the gravity waves test:
\begin{align}
	\phi(z) = \phi_0 \exp \left( \frac{N^2}{g} z \right) \label{eqn:thermal-tracer}
\end{align}
where \(\phi_0 = 288\) and \(N = \SI{0.01}{\per\second}\).  The inlet boundary has fixed values prescribed by equation~\ref{eqn:thermal-tracer} and a zero gradient boundary condition is applied at the outlet.

@@

%% APPENDICES
\begin{appendices}
\input{WellerAppendix}
\end{appendices}
